Efficient Unconstraining Parameter Transforms for Hamiltonian Monte Carlo

### 2022-07-27 dummypaper.tex
updates to simplex part of the paper

%A simplex is a set of non negative vectors that sum up to a constant $a$, this can be written as:
%\begin{equation*}
%\Delta=\left\{\mathbf{x} \in \mathbb{R}^{n} \mid\left\langle\mathbf{x}, \mathbf{1}_{n}\right\rangle=a, \mathbf{x} \geq 0\right\}
%\end{equation*}
%
%In computational works, we generally care about the unit simplex, this is the special case where $ a = 1$. Geometrically, an \textit{n-simplex} is the convex closure of $n+1$ affinely independent set of points $x_0, x_1, \dots x_n$. It is the simplest polytope for $n$-dimensions (in this case we will be looking at a simplex as a convex polytope, not abstract). An \textit{n-simplex} is be represented in the barycentric coordinate system with weighted vertices, as follows:
%\begin{equation*}
%\Delta = \left\{x_{0} u_{0}+\cdots+x_{n} u_{n} \mid \sum_{i=0}^{n} x_{i}=1\\
% \text{ and } x_{i} \geq 0 \text { for } i=0, \ldots, k\right\}
%\end{equation*}
%
%With this representation, taking $u_{i}=1$, the $n$ unit simplex can be written as: 
%\begin{equation*}
%\left\{x \in \mathbb{R}^{n}: x_{0}+\cdots+x_{n-1}=1, x_{i} \geq 0 \text { for } i=0, \ldots, k-1\right\}
%\end{equation*}
%
%A \textit{unit simplex} on $R^n$ is an $n - 1$ dimensional simplex with unit vectors as vertices. Notice the unit weights, this set only has $n-1$ affinely independent points.
%
%Sum to one constraints are useful in probabilistic inference, often used for modeling categorical responses. The rest of the section describes various transforms used to map a simplex to an unconstrained space ($R^n$).\\
%\\

A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
values that sums to one.  Simplexes are useful for representations of multinomial probabilities
(e.g., probabilities of categories in a classification problem). The set of unit $N$-simplexes is conventionally denoted
\[
  \Delta^N = \setcomp{x \in \mathbb{R}_{\ge 0}^{N + 1}}{ \sum_{i=1}^{N} x_i = 1}
\]
Geometrically, an $N$-simplex is the convex closure of $N+1$ points
that are 1 in one coordinate and 0 elsewhere.  For example, the
3-simplex is the complex closure of
$\begin{bmatrix}1 & 0 & 0 \end{bmatrix},
\begin{bmatrix} 0 & 1 & 0 \end{bmatrix}$,
and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$. As such, there are only $N$ degrees of
freedom, because if $x$ is an $N$-simplex, then
\[
  x_N = 1 - (x_1 + x_2 + \cdots + x_{N-1}).
\]
We will use $\Delta^N_-$ to denote $N-1$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
\subsection{StickBreaking Transform}

The StickBreaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}. It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$. Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  R^{N-1}$, for $1 \leq i \leq N$ as:	
\[
y_i
= \mathrm{logit}(z_i) - \mbox{log}\left(\frac{1}{N-i}
   \right) 
   \]

@@

The inverse transform $ f^{-1} \colon R^{N-1} \xrightarrow{\makebox[0.4cm]{$\sim$}}  \Delta^{N-1}$ is defined as:

@@

An $N$ dimensional unit simplex $\Delta^{N-1}$ has $N-1$ degrees of freedom (Notice that this inverse transform only maps the first $N-1$ elements, the last element of the simplex $x_{N} = 1 - \sum_1^{N-1}{x_i}$). The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix, so for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
\begin{align*}
\mathbf{J}_{i, i} &= \frac{\partial x_i}{\partial y_i}
=
\frac{\partial x_i}{\partial z_i} \,
\frac{\partial z_i}{\partial y_i}\\
\mathbf{J}_{i, i} &= \left(
  1 - \sum_{k' = 1}^{k-1} x_{k'}
   \right) z_k (1 - z_k),\\
   \abs{\, det \, \textbf{J} \,} &= \prod_{i=1}^{N-1} \textbf{J}_{i,i}
\end{align*}

The correction term $p_Y(y) = p_X(f^{-1}(y))\,
\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$
\subsection{Additive log ratio transform}

The unconstraining transform for the identified softmax is known as
the additive log ratio (ALR) transform
\cite{aitchison1982statistical}, which is a bijection
$\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
$x \in \Delta^{N-1}$ by
\[
  \textrm{alr}(x)
  = \begin{bmatrix}\displaystyle
    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
  \end{bmatrix}
\]
The inverse additive log ratio transform maps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{alr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,
\[
  \textrm{softmax}(u) = \frac{e^u}{\sum{e^u}}
\]
Again, the change of variables adjustment is calculated for $N-1$ dimensions.  
\[
\begin{bmatrix}
    \textrm{alr}^{-1}(y)_1
    & \cdots &
    \textrm{alr}^{-1}(y)_{N-1}
    \end{bmatrix} = \frac{e^y}{\sum e^y + 1}
\]
From a density $p_X(x)$ defined over $x \in \Delta^{N-1}$,
we can obtain a density over unconstrained parameters $y \in
\mathbb{R}^{N-1}$ by applying the inverse ALR transform and adjusting
for the change of variables
\[
  p_Y(y) = p_X(\textrm{alr}^{-1}(y)) \absdet{J_{f^{-1}}(y)}
\]
where $J_{f^{-1}}$ is the $N-1$ dimensional Jacobian matrix. To calculate the determinant of the Jacobian of the inverse transform,
we start by noting that $alr^{-1}(y) = f \circ g$, where
$f(y) = e^{ \circ y}$ (elementwise exponential), and $g(z)=\frac{z}{\sum z + 1}$. Here $z = e^{\circ y} \in R_{+}^{N-1}$, at this point we evaluate the Jacobian as:
\[
  \absdet{J_{alr^{-1}}(y)}
  = \absdet{J_{f(y)}} \absdet{J_{g(z)}}
\]
Note that $J_{f(z)}$ only consists of diagonal entries $e^{y_{i}}$
\[
  \absdet{J_{f(y)}} = \prod_i^n e^{y_i}
\]
For $g$ we begin with evaluating $ J_{g} = \frac{\partial g(z)}{\partial z}$. Note that $\mathbf{1}$ is a vector of ones.
\[
 J_g = \frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I}_{N-1}
  - \left(\frac{1}{1 + \sum e_{i=1}^{N} z_i^2} x \right)
  \mathbf{1}_{N-1}
\]
Using the matrix determinant lemma,\footnote{The matrix determinant lemma
  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
    \textrm{det}(A).\]}
\begin{eqnarray*}
  |J_{g}|
  & = &
  \left(
    1
    + \mathbf{1}
    \left(\frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I} \right)^{-1}
    \frac{-z}{(1 + \sum e_{i=1}^{N} z_i)^2}
    \right)
    \abs{\frac{1}{1 + \sum e_{i=1}^{N} z_i} \mathbb{I}}
  \\[6pt]

  & = & \left( \frac{1}{1 + \sum e_{i=1}^{N} z_i} \right)^N.
\end{eqnarray*} 
Finally,
\[
  \absdet{J_{alr^{-1}}(y)}
  \ = \
 \prod_i^n e^{y_i}
  \, \left( \frac{1}{1 + \sum e_{i=1}^{N}y_i} \right)^N.
\]
Putting $J_{alr^{-1}}$ in the change of variables adjustment, we get
\[
  p_Y(y)
  = p_X(\textrm{alr}^{-1}(y))
  \, \prod_{i=1}^{N} e^y_i)
  \left( \frac{1}{1 + \sum_{i=1}^{N} e^y_i}\right)^N
\]

%\subsection{Softmax Transform}
%
%The softmax function can be understood from Multinomial Logistic
%Regression employed for predicting probabilities of a Categorically
%distributed variable. Geometrically it maps $R^K$ to the boundary of a
%unit K-simplex(it is simply the convex hull of $k+1$ affinely
%independent points in $R^K$. Essentially it transforms a vector of
%size K to another vector of size K where the outputs sum to 1. It is
%worth noting that the mapping is actually from $R^K$ to $R^{K-1}$, so
%when a vector of size $K$ is transformed the $K_th$ vector is simply
%$1$- sumof k-1 vectors

\subsection{Simplex softmax parameterization}

We define the transformation
$\phi: \mathbb{R}^{n-1} \to \Delta^n_-: y \mapsto x_-$, where
$x=\begin{pmatrix}x_- \\ \frac{1}{r}\end{pmatrix} \in \Delta^n$,
$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$, and
$r = 1 + \sum_{i=1}^{n-1} e^{y_i}$.

First we compute the scalar derivatives:
\[
\begin{aligned}
  \frac{\mathrm{d} r}{\mathrm{d} y_j}
  &= e^{y_j} = r x_j\\
  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i}
  \frac{\mathrm{d} r}{\mathrm{d} y_j}
  = \delta_{ij} x_i - x_i x_j, \quad 1 \le i \le n-1
\end{aligned}
\]
where
$\delta_{ij} = \begin{cases} 1 &\text{if } i = j \\ 0
  &\text{otherwise}\end{cases}$ is the Kronecker delta.

If $\mathrm{diag}(x)$ is the diagonal matrix whose diagonal are the
elements of $x$, then the Jacobian is
\[
  J = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top) \operatorname{diag}(x_-),
\]
where $\boldsymbol{1}_n$ is the $n$-vector of ones.

Using Sylvester's determinant theorem,
$|I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = 1 -
\boldsymbol{1}_{n-1}^\top x_- = 1 - \sum_{i=1}^{n-1} x_i = x_n$, so
$$\mathrm{correction} = |J| = x_n \prod_{i=1}^{n-1} x_i = \prod_{i=1}^{n} x_i = \exp\left(\sum_{i=1}^{n-1} y_i\right) \left(1 + \sum_{i=1}^{n-1} e^{y_i}\right)^{-n}$$

\subsection{Simplex softmax parameterization through parameter expansion}

We define the transformation
$\phi: \mathbb{R}^n \to \Delta^{n-1} \times \mathbb{R}_{>0}: y \mapsto
(x_-, r)$, where $r = \sum_{i=1}^n e^{y_i}$ and
$x_i = \frac{1}{r} e^{y_i}$ for $1 \le i \le n-1$..

First we compute the scalar derivatives:
\[
\begin{aligned}
  \frac{\mathrm{d} r}{\mathrm{d} y_j}
  &= e^{y_j} = r x_j
  \\
  \frac{\mathrm{d} x_i}{\mathrm{d} y_j}
  &= \delta_{ij} \frac{1}{r} e^{y_i} - \frac{1}{r^2} e^{y_i} \frac{\mathrm{d} r}{\mathrm{d} y_j} = \delta_{ij} x_i - x_i x_j,
\end{aligned}
\]
which corresponds to the Jacobian
\[
  J = \begin{pmatrix}I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top & -x_- \\
    r \boldsymbol{1}_{n-1}^\top & r \end{pmatrix} \mathrm{diag}(x).
\]

For invertible $A$, the determinant of the block matrix
$\begin{pmatrix}A & B \\ C & D\end{pmatrix}$ is $|A| |D-CA^{-1}B|$.  A
square matrix is invertible iff its determinant is non-yero.  From the
previous section,
\[
  |I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = x_n > 0,
\]
so the determinant of the Jacobian is
\[
  |J| = x_n \left|r + r \boldsymbol{1}_{n-1}^\top (I_{n-1} - x_-
    \boldsymbol{1}_{n-1}^\top)^{-1} x_-\right|
  \prod_{i=1}^n x_i.
\]

Let $w = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top)^{-1} x_-$. Then,
\[
\begin{aligned}
    w - x_- \sum_{i=1}^{n-1} w_i &= x_-\\
    w &= x_- \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
    \sum_{i=1}^{n-1} w_i &= \sum_{i=1}^{n-1} \left( x_- (1 - \sum_{i=1}^{n-1} w_i) \right) = \left(\sum_{i=1}^{n-1} x_i \right) \left(1 - \sum_{i=1}^{n-1} w_i\right) = (1 - x_n)  \left(1 - \sum_{i=1}^{n-1} w_i\right)\\
    \sum_{i=1}^{n-1} w_i &= \frac{1 - x_n}{x_n} = \frac{1}{x_n} - 1\\
    w &= x_- \left(1 - \left(\frac{1}{x_n} - 1\right)\right) = \frac{1}{x_n} x_-
  \end{aligned}
\]

Then
\[
  |J| = x_n r \left|1 + \frac{1}{x_n}\sum_{i=1}^{n-1} x_i\right|
  \prod_{i=1}^n x_i = r \prod_{i=1}^n x_i
\]

To keep the target distribution proper, we must select a prior
distribution $\pi(r)$ for $r$.  If we choose $r \sim \chi_n$, then the
product of the correction and the density of the prior for $r$ is
proportional to

\[
  \mathrm{correction}
  = \pi(r) |J| = r^n e^{-r^2/2} \prod_{i=1}^n x_i
  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{2}\left(\sum_{i=1}^n
      e^{y_i}\right)^2\right).
\]

Alternatively, if we choose $r \sim \mathrm{Gamma}(n, 1)$, then
\[
  \mathrm{correction} = \pi(r) |J| = r^n e^{-r} \prod_{i=1}^n x_i =
  \exp\left(\sum_{i=1}^n y_i - \sum_{i=1}^n e^{y_i}\right).
\]
This latter correction is equivalent to the sampling procedure from
the Dirichlet distribution with $\alpha_i=1$, where
$z_i \sim \mathrm{Exponential}(1)$ and
$y = \frac{z}{\sum_{i=1}^n z_i}$.

Both of these corrections can be captured with the generalization
\[
  \mathrm{correction}
  = \pi(r) |J|
  = r^n e^{-r^p/p} \prod_{i=1}^n x_i
  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{p} \left(\sum_{i=1}^n e^{y_i}\right)^p\right),
\]
for $p > 0$, which corresponds to $r \sim \text{Generalized-Gamma}(1, n, p)$.

@@

\begin{figure}
    \centering
    \includegraphics[width=1.2\textwidth]{figures/ess.png}
    \caption{Effective Sample Size vs Total LeapFrog Steps}
    \label{fig:ess}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1.2\textwidth]{figures/rmse.png}
    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps}
    \label{fig:rmse}
\end{figure}

@@

>>>>

A unit $N$-simplex is an $N + 1$-dimensional vector of non-negative
values that sums to one.  Simplexes are useful for representations of multinomial probabilities
(e.g., probabilities of categories in a classification problem). The set of unit $N$-simplexes is conventionally denoted
\[
  \Delta^N = \setcomp{x \in \mathbb{R}_{\ge 0}^{N + 1}}{ \sum_{i=1}^{N} x_i = 1}
\]
Geometrically, an $N$-simplex is the convex closure of $N+1$ points
that are 1 in one coordinate and 0 elsewhere.  For example, the
3-simplex is the complex closure of
$\begin{bmatrix}1 & 0 & 0 \end{bmatrix},
\begin{bmatrix} 0 & 1 & 0 \end{bmatrix}$,
and $\begin{bmatrix} 0 & 0 & 1 \end{bmatrix}$. As such, there are only $N$ degrees of
freedom, because if $x$ is an $N$-simplex, then
\[
  x_N = 1 - (x_1 + x_2 + \cdots + x_{N-1}).
\]
%n not n-1
We will use $\Delta^N_-$ to denote $N-1$ elements of a simplex, this is sufficient to uniquely determine $\Delta^N$.
\subsection{Stick-Breaking Transform}

The Stick-Breaking transform carries forward the intuition in the stick-breaking construction for Dirichlet \cite{sethuraman1994constructive}. It is a process of recursively breaking a piece $x_i$ from a stick of unit length, where the leftover stick in the $i^{th}$ iteration is $ 1 - \sum_{1}^{i}x$. Let $y = f(x)$, then we define the stick-breaking mapping $ f \colon \Delta^{N-1} \to  R^{N-1}$, for $1 \leq i \leq N$ as:	
\[
y_i
= \mathrm{logit}(z_i) - \mbox{log}\left(\frac{1}{N-i}
   \right) 
   \]

@@

The inverse transform $ f^{-1} \colon R^{N-1} \to \Delta^{N-1}$ is defined as:

@@

The determinant of the Jacobian
\[
   \abs{\mathbf{J}} = \prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right)
\]
\subsection{Additive log ratio transform}
The unconstraining transform for the identified softmax is known as
the additive log ratio (ALR) transform
\cite{aitchison1982statistical}, which is a bijection
$\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
$x \in \Delta^{N-1}$ by
\[
  \textrm{alr}(x)
  = \begin{bmatrix}\displaystyle
    \log \frac{x_1}{x_N} \cdots \log \frac{x_{N-1}}{x_N}
  \end{bmatrix}
\]
The inverse additive log ratio transform maps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{alr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,

\[
  \textrm{softmax}(u) = \frac{exp(u)}{\sum exp(u)}
\]

\[
\textrm{Here, } \abs{J} = \prod \exp(y)
  \, \left( \frac{1}{1 + \sum(\exp(y))} \right)^N
\]
\subsection{Augmented-Softmax Transform}

We define the transformation
$\phi: \mathbb{R}^n \to \Delta^{n-1} \times \mathbb{R}_{>0}: y \mapsto
(x_-, r)$, where  $\Delta_{-}^N$ and $x_{-}$  denote $N-1$ elements of the simplex and $x$, respectively. Here $r = \sum_{i=1}^n exp(y_i)$,
$x_i = \frac{1}{r} exp(y_i)$ for $i \in [1, N-1]$ and $\delta_{ij}$ is the Kronecker delta function. If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $n$-vector of ones.\[
  J = (I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top) \operatorname{diag}(x_-),
\]

Using Sylvester's determinant theorem,
$|I_{n-1} - x_- \boldsymbol{1}_{n-1}^\top| = 1 -
\boldsymbol{1}_{n-1}^\top x_- = 1 - \sum_{i=1}^{n-1} x_i = x_n$, so
$$ |J| = x_n \prod_{i=1}^{n-1} x_i = \prod_{i=1}^{n} x_i = \exp\left(\sum_{i=1}^{n-1} y_i\right) \left(1 + \sum_{i=1}^{n-1} e^{y_i}\right)^{-n}$$

@@

\begin{figure}
    \centering
    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_density.png}
    \caption{Effective Sample Size/Total LeapFrog Steps}
    \label{fig:ess_density}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1.2\textwidth]{figures/simplex/ess_cdf.png}
    \caption{Effective Sample Size/Total LeapFrog Steps}
    \label{fig:ess_cdf}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1.2\textwidth]{figures/simplex/rmse.png}
    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps}
    \label{fig:rmse}
\end{figure}

@@

\appendix
\section{Jacobian Determinants}
\subsection{Stickbreaking Transform}
The Jacobian matrix for $f^{-1}$ is a lower-triangular diagonal matrix, so for the change of variables we evaluate $\mathbf{J}_{i, i}$ where $i \in 1:N-1$.
\begin{align*}
\mathbf{J}_{i, i} &= \frac{\partial x_i}{\partial y_i}
=
\frac{\partial x_i}{\partial z_i} \,
\frac{\partial z_i}{\partial y_i}\\
\mathbf{J}_{i, i} &= \left(
  1 - \sum_{k' = 1}^{k-1} x_{k'}
   \right) z_k (1 - z_k),\\
   \abs{\textbf{J}} &= \prod_{i=1}^{N-1} \textbf{J}_{i,i}
\end{align*}

The change of variables adjustment $p_Y(y) = p_X(f^{-1}(y))\,
\prod_{i=1}^{N-1}z_i\,(1 - z_i)\left(1 - \sum_{i'=1}^{i-1} x_{i'}\right).$

\subsection{Additive Log ratio transform}
To calculate the determinant of the Jacobian of the inverse transform,
we start by noting that $s = \textrm{exp} \circ \textrm{norm}$, where
$\textrm{exp}$ is the elementwise exponential function and
\textrm{norm} is defined by
\[
  \textrm{norm}(z) = \frac{z}{\textrm{sum}(z) + 1}.
\]
As such, the resulting Jacobian determinant is the product of the
Jacobian determinants of the component functions,
\[
  \absdet{J_s(y)}
  = \absdet{J_{\textrm{exp}}(y)} \absdet{J_{\textrm{norm}}(z)},
\]
where $z = \textrm{exp}(y)$.  The Jacobian for the exponential
function is diagonal, so the determinant is the product of the
diagonal of the Jacobian, which for $y \in \mathbb{R}^{N-1}$ is
\[
  \absdet{J_{\textrm{exp}}(y)} = \textrm{prod}(\exp(y)).
\]
As above, let $z = \exp(y) \in (0, \inf)^{N-1}$.  We can differentiate
$\textrm{norm}$ to derive the Jacobian,
\[
  J_{\textrm{norm}}
  = \frac{1}{1 + \textrm{sum}(z)} \mathbb{I}_{N-1}
  - \left(\frac{1}{(1 + \textrm{sum}(z))^2} \beta \right)
  \textrm{vector}_{N-1}(1)^{\top},
\]
where $\mathbb{I}_{N-1}$ is the $(N - 1) \times (N - 1)$ unit matrix and
$\textrm{vector}_{N-1}(1)$ is the $N - 1$-vector with values 1.  Using
the matrix determinant lemma,\footnote{The matrix determinant lemma
  is \[\textrm{det}(A + u v^{\top}) = (1 + v^{\top} A^{-1} u)
    \textrm{det}(A).\]}
we have
\begin{eqnarray*}
  \textrm{absdet}(J_{\textrm{norm}}(z))
  & = &
  \left(
    1
    + \textrm{vector}_{N-1}(1)^{\top}
    \left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I} \right)^{-1}
    \frac{-z}{(1 + \textrm{sum}(z))^2}
    \right)
    \ \textrm{det}\left(\frac{1}{1 + \textrm{sum}(z)} \mathbb{I}
        \right)
  \\[6pt]
  & = &
  \left(
    1 
    + \textrm{sum}\left( \frac{-(1 + \textrm{sum}(z)) z}{(1 +
        \textrm{sum}(z))^2} \right)
  \right)
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = &
        \left(1 + \textrm{sum}\left(\frac{-z}{1 + \textrm{sum}(z)} \right)\right)        
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( 1 - \textrm{sum}(\textrm{norm}(z)) \right) 
        \ \left( \frac{1}{1 + \textrm{sum}(z)} \right)^{N-1}
  \\[6pt]
  & = & \left( \frac{1}{1 + \textrm{sum}(z)} \right)^N.
\end{eqnarray*}
Thus the entire absolute determinant of the Jacobian is defined by the
product, 
\[
  \absdet{J_s(y)}
  \ = \
  \textrm{prod}(\exp(y))
  \, \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N.
\]
and our final expression for densities for unconstrained $y \in
\mathbb{R}^{N-1}$ is
\[
  p_Y(y)
  = p_X(\textrm{alr}^{-1}(y))
  \, \textrm{prod}(\exp(y))
  \left( \frac{1}{1 + \textrm{sum}(\exp(y))} \right)^N
\]
